---
title: "Danceability of US and UK Top Artists"
author: 
  - Aman Rana
  - Daniel Du
  - Michelle Ji
thanks: "Code and data are available at: [https://github.com/danield424/Danceability-of-US-and-UK-Top-Artists](https://github.com/danield424/Danceability-of-US-and-UK-Top-Artists)."
date: today
date-format: long
abstract: "Album data from the top three artists in the United States and the United Kingdom is analyzed in this report. The focus of the album data analysis is on key music features that can be of importance to artists when marketing their music and gauging their audience. Data is used from @billboard and @musicmetricsvault. Overall, we find that the top three artists in the United Kingdom have higher danceability levels than the top three artists in the United States, potentially indicating the music listeners in the United Kingdom may value higher danceability."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
```


# Introduction

The music industry continues to increase significance in everyday life, from listening to music on a daily commute to school or work to music played in stores. Culture and region also plays a role in what artists and songs are popular, as seen with various Billboard 100 lists from different countries. This paper examines the difference between music listening in the United States and the United Kingdom, focusing on danceability as the main component for music listening habits. Our findings highlight key features of music that artists could use to see what is popular with their listeners the most, depending if they are a more popular artist in a specific region. The paper is broken down into various sections. @sec-data explores the data, highlighting certain key aspects of the danceability of artists’ albums over the years. @sec-results introduces observations and potential conclusions from the data analysis.


# Data {#sec-data}

We use @citeR and @tidyverse. We pull @spotify artist audio features data using @spotifyr a wrapper around Spotify’s API. We find the top three US artists from @billboard and the top three artists in the UK from @musicmetricsvault. We drop Chappell Roan as a top three candidate because of her short album history. The artists’ danceability are averaged out to create country-level danceability averages over time.

```{r, include=FALSE}
# Read in data 
#| echo: false
sabrina_carpenter <- read_rds("../data/01-raw_data/sabrina_carpenter.rds")
taylor_swift <- read_rds("../data/01-raw_data/taylor_swift.rds")
billie_eilish <- read_rds("../data/01-raw_data/billie_eilish.rds")
dua_lipa <- read_rds("../data/01-raw_data/dua_lipa.rds")
ed_sheeran <- read_rds("../data/01-raw_data/ed_sheeran.rds")
adele <- read_rds("../data/01-raw_data/adele.rds")
```



```{r}
#| echo: false
#| label: fig-1
#| warning: false
#| fig-cap: The average danceability over time of the songs of the current top 3 US Artists Billie Eilish, Taylor Swift, and Sabrina Carpenter.
#| out-width: 90%

rbind(sabrina_carpenter, billie_eilish, taylor_swift) |>
  drop_na() |>
  select(artist_name, album_release_date, danceability) |>
  mutate(album_release_date = ymd(album_release_date)) |>
  group_by(album_release_date) |>
  summarize(danceability = mean(danceability)) |> 
  ggplot(aes(x = album_release_date, y = danceability)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  theme_minimal() +
  labs(
    x = "Album release date",
    y = "Average Danceability",
    title = "Average Danceability Trend of Current Top 3 US Artists"
  ) +
  theme(legend.position = "none")
```







```{r}
#| echo: false
#| label: fig-2
#| warning: false
#| fig-cap: The average danceability over time of the songs of the current top 3 UK Artists Ed Sheeran, Dua Lipa, and Adele.
#| out-width: 87%

rbind(ed_sheeran, dua_lipa, adele) |>
  drop_na() |>
  select(artist_name, album_release_date, danceability) |>
  mutate(album_release_date = ymd(album_release_date)) |>
  group_by(album_release_date) |>
  summarize(danceability = mean(danceability)) |> 
  ggplot(aes(x = album_release_date, y = danceability)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  theme_minimal() +
  labs(
    x = "Album release date",
    y = "Average Danceability",
    title = "Average Danceability Trend of Current Top 3 UK Artists"
  ) +
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| label: fig-3
#| warning: false
#| fig-cap: The average danceability over time of the combined songs of both the current top 3 US and UK Artists, compared.
#| fig-align: "center"
#| out-width: 87%

# average danceability for the US artists
USTop3 <- rbind(sabrina_carpenter, billie_eilish, taylor_swift) |>
  drop_na() |>
  select(artist_name, album_release_date, danceability) |>
  mutate(album_release_date = ymd(album_release_date)) |>
  group_by(album_release_date) |>
  summarize(danceability = mean(danceability)) |>
  mutate(group = "US Top 3")

# average danceability for the UK artists
UKTop3 <- rbind(ed_sheeran, dua_lipa, adele) |>
  drop_na() |>
  select(artist_name, album_release_date, danceability) |>
  mutate(album_release_date = ymd(album_release_date)) |>
  group_by(album_release_date) |>
  summarize(danceability = mean(danceability)) |>
  mutate(group = "UK Top 3")

# Combine the two groups
combined_groups <- bind_rows(USTop3, UKTop3)

# Plot the combined data with a single trendline for each group
ggplot(combined_groups, aes(x = album_release_date, y = danceability, color = group)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  theme_minimal() +
  labs(
    x = "Album release date",
    y = "Average Danceability",
    color = "Group",
    title = "Danceability Trend of US and UK Artists"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```




\newpage

# Results {#sec-results}

We find that the top three artists from the United Kingdom, which include Dua Lipa, Ed Sheeran, and Adele, tend to have higher danceability levels than the top three artists from the United States, which include Sabrina Carpenter, Taylor Swift, and Billie Eilish. @fig-1 details the average danceability across the top three United States artists. @fig-2 shows the average danceability across the top three United Kingdom artists. @fig-3 highlights the average danceability of top three artists in both the US and the UK. From this graph, we can see that the UK top three artists have higher average danceability levels than the US. 


# References


